Explore the SingleR results

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
Update ordering for consensus dotplot

total_cells_df$y_label <- factor(total_cells_df$y_label, levels =  total_cells_df$y_label)
nbatlas_bar_order <- total_cells_df$label_recoded
```

@@

@@

This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.

@@

# get total number of cells per final annotation group and set up y_label
total_cells_df <- singler_recoded_df |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::arrange(desc(total_cells)) |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))

singler_order <- total_cells_df$y_label
total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
```

```{r}
# prepare markers for dotplot
dotplot_consensus_markers_df <- consensus_markers_df |>
  # consider only matching labels
  dplyr::filter(validation_group_annotation %in% singler_recoded_df$label_recoded) |>
  dplyr::select(
    marker_gene_label = validation_group_annotation,
    ensembl_gene_id,
    gene_symbol
  )

# get the bar order
consensus_bar_order <- total_cells_df |>
  dplyr::select(label_recoded, y_label) |>
  # inner!!!!
  dplyr::inner_join(
    dotplot_consensus_markers_df, 
    by = c("label_recoded" = "marker_gene_label")
  ) |>
  dplyr::arrange(y_label) |>
  dplyr::pull(label_recoded) |>
  unique()
```

>>>>

total_cells_df$y_label <- factor(total_cells_df$y_label, levels = rev(total_cells_df$y_label))
nbatlas_bar_order <- total_cells_df$label_recoded
```

@@

Importantly, these marker genes are not necessarily canonical marker genes for the normal/immune cell types present, but were empirically-determined based on the `NBAtlas` expression patterns.
There may therefore be overlapping marker genes between categories, and the given marker genes may not be the "most" precise for the cell types if analyzed independently.

@@

This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.
Harmonized `SingleR` labels with direct matches are shown on the top of the y-axis (`endothelial cell` through `plasma cell`), and the remaining `SingleR` labels are below.

@@

# to match up consensus and NBAtlas for ordering
direct_celltype_matches <- c(
  "B cell",
  "T cell",
  "myeloid",
  "macrophage", 
  "monocyte",
  "fibroblast",
  "dendritic cell",
  "endothelial cell",
  "plasma cell",
  "natural killer cell"
)

total_cells_df <- singler_recoded_df |>
  dplyr::count(label_recoded, name = "total_cells") |>
  # labels that appear in both consensus+nbatlas should appear first
  dplyr::mutate(in_both = label_recoded %in% direct_celltype_matches) |>
  dplyr::group_by(in_both) |>
  dplyr::arrange(total_cells, .by_group = TRUE) |>
  dplyr::ungroup() |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))
total_cells_df$y_label <- factor(total_cells_df$y_label, levels = total_cells_df$y_label)

# get the annotation bar order
consensus_bar_order <- total_cells_df |>
  dplyr::filter(in_both) |>
  dplyr::pull(label_recoded) |>
  rev()
```

```{r}
# prepare markers for dotplot
dotplot_consensus_markers_df <- consensus_markers_df |>
  # consider only matching labels
  dplyr::filter(validation_group_annotation %in% singler_recoded_df$label_recoded) |>
  dplyr::select(
    marker_gene_label = validation_group_annotation,
    ensembl_gene_id,
    gene_symbol
  )
```
